---
title: Multi-level labels with ggplot2
author: Dmitrijs Kass
date: '2019-06-30'
slug: multi-level-labels-with-ggplot2
categories:
  - data-viz
tags: []
output:
  blogdown::html_page:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = 'center', fig.height = 3, fig.width = 6)
```

Recently I faced a necessity to create **multi-level labels with ggplot2** and had no idea how to do it. Multi-level labels imply some sort of hierarchical strusture in data. For example, survey questions may be assigned into groups; and dates on the timeline may be assigned into years. A 15-minute Google-fu provided me with various solutions described on Stack Overflow that worked with varying success for different types of charts. An important aspect is whether data points between groups should be connected. The bar chart (plot A below) is an example where data points from different groups are not connected and a line chart (plot B below) is an example where data points are connected both within and between groups:

```{r echo = FALSE, fig.height = 5, fig.width = 6, fig.align='center'}
library(dplyr)
library(ggplot2)
library(stringr)

# Bar chart ----

TIMES <- c(2, 4, 3)
set.seed(4)
data <- tibble(group = c(rep(paste("Group", 1:3), times = TIMES)),
               question = paste0("Q", 1:sum(TIMES)),
               proportion = runif(sum(TIMES)))

p_bars <- data %>% 
  ggplot(aes(x = question, y = proportion)) +
  geom_col() +
  facet_grid(~group, scales = "free_x", switch = "x", space = "free_x") +
  theme(strip.placement = "outside", 
        strip.background = element_rect(fill = "white"),
        axis.title = element_blank()) +
  labs(tag = "A")

# Line chart ----

N <- 16
set.seed(4)
data <- tibble(date = seq(as.Date("2018-11-01"), by = "1 month", length.out = N),
               sales = 100 + rnorm(n = N, mean = 10, sd = 10))

format_dates <- function(x) {
  
  years <- lubridate::year(x)
  months <- str_sub(str_to_upper(strftime(x, format = "%B")), start = 1, end = 1)
  
  if_else(is.na(lag(years)) | lag(years) != years, 
          true = paste(months, years, sep = "\n"), 
          false = months)
  
}

p_line <- data %>% 
  ggplot(aes(x = date, y = sales)) +
  geom_line() +
  scale_x_date(name = NULL,
               breaks = "1 month",
               minor_breaks = NULL,
               expand = expand_scale(add = 15),
               labels = format_dates) +
  theme(axis.text.x = element_text(hjust = 0)) +
  labs(tag = "B",
       y = NULL)

gridExtra::grid.arrange(p_bars, p_line, nrow = 2)
```

Below I discuss alternative solution to multi-level labels for these two charts. I deliberately keep them simple, like no customization of y axis labels and colours.

# Bar chart

## Alternative solutions

Bar chart is a good example of a chart where data points between groups don't need to be connected. The most popular solutions employ faceting with either `facet_grid()` (for example, 
[here](https://stackoverflow.com/questions/48552671/ggplot2-show-category-and-sub-category-for-x-axis-labels/48555862#48555862)) 
or `facet_wrap()` (for example, 
[here](https://stackoverflow.com/questions/44247239/grouping-on-the-x-axis-in-ggplot2/44249048#44249048)). 
The latter does not have an argument `space` that lets the width of facets vary, which in its turn forces all bars to have the same width. Therefore, if a number of categories differs among groups then `facet_grid()` should be preferred. I really liked this approach and demonstrate it below in steps.


## Example with faceting

First, let's simulate the data for the chart. Let it be a result of a hypothetical survey with `r sum(TIMES)` questions labeled from Q1 to `r paste0("Q", sum(TIMES))` and combined into `r length(TIMES)` groups. These will substitute x axis labels. The y axis values are drawn from a uniform distribution and represent a proportion of respondents that correctly answered corresponding questions.

```{r}
# For `tibble` and pipe `%>%`.
library(dplyr)
# For reproducibility.
set.seed(4) 
# Counts of elements in each of three group.
TIMES <- c(2, 4, 3)
# Simulate data.
data <- tibble(group = c(rep(paste("Group", 1:3), times = TIMES)),
               question = paste0("Q", 1:sum(TIMES)),
               proportion = runif(sum(TIMES)))
# Print data.
data
```

The faceting approach starts with a simple bar chart that is turned into [small multiples](https://en.wikipedia.org/wiki/Small_multiple) bar chart with `facet_grid()`:

```{r}
library(ggplot2)
data %>% 
  ggplot(aes(x = question, y = proportion)) +
  geom_col() +
  facet_grid(~group)
```

Let's rewrite the code and add three additional arguments for the faceting function:

```{r}
p_bars <- data %>% 
  ggplot(aes(x = question, y = proportion)) +
  geom_col() +
  facet_grid(~group, 
             scales = "free_x", # Let the x axis vary across facets.
             space = "free_x",  # Let the width of facets vary and force all bars to have the same width.
             switch = "x")      # Move the facet labels to the bottom.

p_bars
```

The final step is to customize the non-data components of the chart with the `theme()` function:

```{r}
p_bars + 
  theme(strip.placement = "outside",                      # Place facet labels outside x axis labels.
        strip.background = element_rect(fill = "white"),  # Make facet label background white.
        axis.title = element_blank())                     # Remove x and y axis titles.
```

Done. We may remove the white space between facets by providing a `panel.spacing = unit(0.03, units = "cm")` argument to the `theme()` function. But I don't recommend it because the borders between groups get blurred and readability of the chart suffers.




# Line chart

## Alternative solutions

Faceting approach does not fit when all data points need to be connected as demonstrated in the question 
[here](https://stackoverflow.com/questions/20571306/multi-row-x-axis-labels-in-ggplot-line-chart).

One solution offered 
[here](https://stackoverflow.com/questions/20571306/multi-row-x-axis-labels-in-ggplot-line-chart/20574024#20574024)
is to use annotations. I don't use this approach for two reasons:

1. `coord_cartesian(clip = "off")` can cause unexpected results as it allows drawing of data points anywhere on the plot, including in the plot margins.
2. Fixed horizontal positions for text annotations in `annotate(geom = "text", x = 2.5 + 4 * (0:4))` do not depend on data and other ggplot2 customizations, such as axis limits. This solution may not work for a different data set.

Another solution with grobs 
[here](https://stackoverflow.com/questions/44616530/axis-labels-on-two-lines-with-nested-x-variables-year-below-months/44616739#44616739) appears overly complicated to me. Instead, I offer a different solution, which seems to be simple and generalizable, i.e. does not depend on a particular data set.


## Example with text wrap

Again, we start with simulating data for the line chart. Let it be sales volume of some healthy product during 16 months from November 2018 to February 2020. I deliberately chose this period as it partialy covers three years.

```{r}
# Number of months.
N <- 16
# Reproducibility.
set.seed(4)
# Simulate data.
data <- tibble(date = seq(as.Date("2018-11-01"), by = "1 month", length.out = N),
               sales = 100 + rnorm(n = N, mean = 10, sd = 10))
# Print data.
data
```

The first step is to create a simple line chart:



```{r}
p_line <- data %>% 
  ggplot(aes(x = date, y = sales)) +
  geom_line()

p_line
```

The x axis labels will vary depending of regional settings. My default region is Latvia. Locale can be changed with `Sys.setlocale()`:

```{r}
# Change locale.
Sys.setlocale(category = "LC_ALL", locale = "english")
# Reprint the chart.
p_line
```

The faceting approach is not suitable in this scenario because we need to connect dots between facets:

```{r}
# Faceting approach does not suit this chart.
data %>% 
  ggplot(aes(x = date, y = sales)) +
  geom_line() +
  facet_grid(~lubridate::year(date), scales = "free_x", switch = "x", space = "free_x") + 
  theme(strip.placement = "outside", 
        strip.background = element_rect(fill = "white"), 
        axis.title = element_blank())
```


The simplest approach in my view is to modify x axis labels with a simple function that does three things:

1. Gets the first letter of the month's name. Depending on your locale you may or may not need to capitalize this letter. I would need it if I kept "latvian" locale, so I will keep it inside my function below.
2. Gets the year part from the date.
3. Paste the year behind the first letter of the month using a new line `\n` as a separator, if 
    * it is the first element in a date vector, or
    * the previous date has a different year component.

```{r}
library(stringr)
format_dates <- function(x) {
  
  # First letter of the most, capitalized.
  months <- str_sub(str_to_upper(strftime(x, format = "%B")), start = 1, end = 1)
  # Year as a 4-digit number.
  years <- lubridate::year(x)
  if_else(is.na(lag(years)) | lag(years) != years, 
          true = paste(months, years, sep = "\n"), 
          false = months)
  
}
```


For example:

```{r}
# Test the function of quarterly data.
format_dates(as.Date(c("2018-09-30", "2018-12-31", "2019-03-31", "2019-06-30")))
format_dates(as.Date(c("2018-12-31", "2019-03-31", "2019-06-30", "2019-09-30")))
```

Now simply modify the `p_line` chart by adding `scale_x_date()` with few arguments. Supply the above defined date formatting function to the `labels` argument. As the final touch, left-justify the labels with the `theme()`.

```{r}
p_line + 
  # Customize x axis.
  scale_x_date(date_breaks = "1 month",          # Date labels for each month.
               minor_breaks = NULL,              # No additional labels in-between `date_breaks`.
               expand = expand_scale(add = 15),  # Add 15 days to the x-axis on the left and on the right.
               labels = format_dates) +          # Supply a user-defined date formatting function.
  
  # Customize the non-data components.
  theme(axis.text.x = element_text(hjust = 0),   # Left-justify x-axis labels.
        axis.title = element_blank())            # Remove x and y axis titles.
```

<br>

Any comments or suggestions? I'd be glad to know! Please leave them below, no login required if you check "I'd rather post as a guest".


<!-- ```{r echo = FALSE} -->
<!-- # Super-categories for x-axis (in the end came up with a simple solution with labels). -->
<!-- # -->
<!-- # Faceting: -->
<!-- # https://stackoverflow.com/questions/48552671/ggplot2-show-category-and-sub-category-for-x-axis-labels -->
<!-- # https://stackoverflow.com/questions/44247239/grouping-on-the-x-axis-in-ggplot2 -->
<!-- # -->
<!-- # Annotations: -->
<!-- # https://stackoverflow.com/questions/20571306/multi-row-x-axis-labels-in-ggplot-line-chart -->
<!-- # -->
<!-- # Grobs: -->
<!-- # https://stackoverflow.com/questions/44616530/axis-labels-on-two-lines-with-nested-x-variables-year-below-months/44616739 -->
<!-- # -->
<!-- # https://github.com/tidyverse/ggplot2/issues/1966 -->
<!-- ``` -->

